\documentclass[10pt,a5paper]{book}

\usepackage[oldstyle]{libertine}
\usepackage{lettrine}
\usepackage{xcolor}
\usepackage{verse}
\usepackage{titling}
\usepackage{fancyhdr}
\renewcommand{\headrulewidth}{0pt}


% BLUES
\definecolor{benblue1}{HTML}{2B22C7}
% REDS
\definecolor{benred8}{HTML}{E82C00} %seems good, maybe the best?

\title{Pray Without Ceasing}
\date{October 2013}
\author{Fr Cassian Folsom OSB}

\newenvironment{poem}
  {\begin{itshape}\begin{verse}
  }
  {\end{verse}\end{itshape}
  }

\begin{document}

\maketitle

\pagestyle{empty}

\begin{poem}
\textsc{O lux} beata Trinitas, \\
et principalis Unitas, \\
iam sol recedit igneus, \\
infunde lumen cordibus.

Te mane laudum carmine, \\
te deprecemur vespere: \\
te nostra supplex gloria \\
per cuncta laudet saecula.

Deo Patri sit gloria, \\
eiusque soli Filio, \\
cum Spiritu Paraclito, \\
et nunc, et in perpetuum.
\end{poem}

\newpage

Fr Cassian was invited to St Benedict's Abbey in Still River, Massachusetts in the early October 2013 to give a series of conferences on praying without ceasing. Drawing on Scripture, the Liturgy and the Fathers, Fr Cassian gives us some tips on how to pray better. In this day and age, we want quick fixes and easy answers, but we can only learn to pray better by doing it more frequently and this takes time.

Fr Cassian is the founder and prior of the Monastery of San Benedetto, located at the birthplace of St Benedict in Norcia, Italy. Fr Cassian has remained on faculty at the Pontifical Liturgical Institute in Rome, and since 2010 has been a consultor to the Congregation of Divine Worship and the Discipline of the Sacraments. Fr Cassian became a Benedictine monk at St Meinrad in southern Indiana in 1980 and was ordained a priest in 1984. He earned his doctorate in Sacred Liturgy from Sant'Anselmo in 1989. He has given retreats and talks throughout his life as a monk.

\input{lecture01}

\end{document}